---
title: "Group Project BIOL432"
author: "R&B(ioinformatics)"
date: "06/04/2021"
output: html_document
---

# Project Info

**Date**: April 6th, 2021

**Group members:**

Amanda Zacharias (20043354)

Arjun Augustine (20060723)

Alexander Row (20067626)

Yudong Liu (20020036)

Cameron MacPhail (20012451)



**GitHub Repository Link**: [link](https://github.com/alexrow15/BIOL432.GroupProject)

## Load Required Packages
```{r}
library(dplyr)
library(ggplot2)
```

## Load Dataset
```{r}
LadyBugData <- read.csv("BIOL432.Data.GroupProject.csv")
```

## Visualise Raw Data
```{r}
head(LadyBugData)
```

```{r}
tail(LadyBugData)
```

```{r}
summary(LadyBugData) #view summary for each variable in dataset
<<<<<<< HEAD
```
=======
```


### A plot for demonstration of the custom function
 
made by Cameron
```{r}
tempAbundancePlot <- ggplot(dat, aes(x = DOY, y = Adults)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```


## Making A Custom Function

This function will save a plot in a specified file format. File format options are pdf, png, and jpeg. File will be output to a folder called, "Output".

\n

For example: 

+ input: plot = tempAbundancePlot, file_type = .png, directory = "Folder/Filename.png", width = 5, height = 5 
+ output = a .png image of the graph that is named, "graph"

```{r}
outputFigGenerator <- function(plot_code, file_type = 0,
                               directory="Output/name.png",
                               fig.width = 5,
                               fig.height = 5){
  if (file_type == 1){
    pdf(file = directory, width = fig.width,
            height = fig.height)
    plot_code
  }
  if (file_type == 2){
    png(file = directory, width = fig.width,
            height = fig.height)
    plot_code
  }
  if (file_type == 3){
    jpeg(file = directory, width = fig.width,
            height = fig.height)
    plot_code
  }
  dev.off()
}
```

Run the function

```{r}
outputFigGenerator(plot_code = ggplot(dat, aes(x = DOY, y = Adults)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw(), file_type = 2, directory =  "Output/tempAbundancePlot.png", fig.width = 5, fig.height = 5)
```

```{r}
png(file = "Output/tempAbundancePlot.png")
tempAbundancePlot
dev.off()
```





>>>>>>> 0dec02c71f787801344243d5c59c85ebc9822a4d
