---
title: "Group Project BIOL432"
author: "R&B(ioinformatics)"
date: "06/04/2021"
output: html_document
---

# Project Info

**Date**: April 6th, 2021

**Group members:**

Amanda Zacharias (20043354)

Arjun Augustine (20060723)

Alexander Row (20067626)

Yudong Liu (20020036)

Cameron MacPhail (20012451)

**We seek to address the following questions:**

1. Within sampling years, does native and non-native species abundance change according to the month?

2. Is there seasonal niche partitioning among native species and among non-native species? Is there seasonal niche partitioning between native and non-native species?

3. Is there a change in seasonal partitioning before and after invasive species arrive?


[GitHub Repository Link](https://github.com/alexrow15/BIOL432.GroupProject)

## Load Required Packages
```{r message=F}
library(dplyr)
library(ggplot2)
```

## Load Dataset
```{r}
LadyBugData <- read.csv("BIOL432.Data.GroupProject.csv")
```

## Visualise Raw Data
```{r}
head(LadyBugData)
```

```{r}
tail(LadyBugData)
```

```{r}
summary(LadyBugData) #view summary for each variable in dataset
```



