---
title: "Question_1_code"
author: "Cameron MacPhail"
date: "16/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

###Within sampling years, does native and non-native species abundance change according to the month?

Loading packages
```{r, echo = FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

Importing Data
```{r}
dat <- read.csv("./BIOL432.Data.GroupProject.csv")
head(dat)
```

The dates loaded in correctly, which is good.

## General temporal abundance patterns

So, does abundance vary at all over time?
```{r}
ggplot(dat, aes(x = DOY, y = Adults)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```

Clearly abundance does vary over time, but not necessarily exactly according to month. Let's examine how well this pattern holds for individual species as well as native vs non-native assemblages.

Creating native and non-native groups
```{r}
dat <- dat %>%
  mutate(Species_type = ifelse(SpeciesID == "CSEPT" |  SpeciesID == "HAXY" | SpeciesID == "HVAR" | SpeciesID == "PQUA", "Non-native", "Native"))
```


## Temporal abundance of native vs non-native ladybeetles

Plotting abundance w/ native and non-native groupings
```{r}
ggplot(dat, aes(x = DOY, y = Adults, colour = Species_type)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```

Non-native ladybeetles clearly have higher abundance throughout the entire year in most habitats. There doesn't seem to be strong patterns in forest habitat for native ladybeetles, but this might just be because they have such low abundance. Taking a look at abundance pre-invasion of the various exotic species should help show what's going on. Also, in several habitats there seems to be a decline in non-native ladybeetle abundance and a concurrent increase in native ladybeetle abundance around 2/3 of the way through the year, lending support to our hypothesis of temporal niche partitioning.


## Pre-invasion temporal abundance patterns

1989 - 1993: one non-native ladybeetles
```{r}
dat %>%
  filter(Year == c(1989:1993), TYPE != "forest") %>% #filtering desired years and removing forest data (forest site only started getting sampled in 1993, not enough data for this time frame)
  ggplot(aes(x = DOY, y = Adults, colour = Species_type)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```
Native species are already low in abundance by this point, but seem to do well during the shoulder seasons.

```{r}
```


1994 - 1998: two non-native ladybeetles
```{r}
dat %>%
  filter(Year == c(1994:1998)) %>% #filtering desired years
  ggplot(aes(x = DOY, y = Adults, colour = Species_type)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```

Very interesting change in pattern for annual crops. Graphing by individual species (replace "Species_type" with "SpeciesID") shows that the increase in native abundance of is driven almost entirely by one species.


1999 - 2006: three non-native ladybeetles + soybean aphid outbreaks
```{r}
dat %>%
  filter(Year == c(1999:2006)) %>% #filtering desired years
  ggplot(aes(x = DOY, y = Adults, colour = Species_type)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```

During this time period, a new, exotic ladybeetle species arrived and there were repeated soybean aphid outbreaks. Things get a little weird....

2007 - 2012: four non-native ladybeetles + soybean aphids present
```{r}
dat %>%
  filter(Year == c(2007:2012)) %>% #filtering desired years
  ggplot(aes(x = DOY, y = Adults, colour = Species_type)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  geom_smooth() +
  facet_wrap(~ TYPE) +
  theme_bw()
```

That pattern of mid to late season niche partitioning becomes very apparent in annual cropland. Potentially related to agricultural activities (e.g. harvest)? Digging into the patterns for specific treatment types might explain more.


## Conclusion

Overall, abundance certainly varies over time and we see some evidence for niche partitioning at this stage. Invasion of new species also definitely has an impact on patterns.


## Question 2: Is there seasonal niche partitioning among native species and among non-native species? Is there seasonal niche partitioning between native and non-native species?




